Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2019-10-05 manuscript/manuscript.tex
changes to intro

A long-standing problem in evolutionary genetics is quantifying the roles of
genetic drift and selection in shaping genome-wide allele frequency changes.
Selection can both directly and indirectly affect allele frequencies, with the
indirect effect coming from the action of selection on correlated loci
elsewhere in genome e.g. linked selection (\cite{Maynard_Smith1974-lc},
\cite{Charlesworth1993-gb,Nordborg1996-nq}; see \cite{Barton2000-zg} for a
review). Previous work on this question has mostly focused on teasing apart
the impacts of drift and selection on genome-wide diversity using population
samples from a single contemporary timepoint, often by modeling the correlation
between regional recombination rate, gene density, and diversity created in the
presence of linked selection \parencite{Begun1992-ey,Elyashiv2016-vt}. This
approach has shown linked selection has a major role in shaping patterns of
genome-wide diversity across the genomes of some of organisms
\parencite{Begun2007-bg,Beissinger2016-cm,Sattath2011-dr,Williamson2014-oy,Andersen2012-bj,Cutter2010-gi},
and has allowed us to quantify the relative influence of positive selection
(hitchhiking) and negative selection (background selection;
\cite{Nordborg2005-dc,McVicker2009-ax,Hernandez2011-gs,Elyashiv2016-vt}.
However, we lack an understanding of how genome-wide linked selection acts over
time.

There are numerous examples of rapid phenotypic adaptation
\parencite{Grant2011-wk,Grant2006-hj,Reznick1997-mh,Franks2007-dr} and rapid
genomic evolution in asexual populations
\parencite{Good2017-om,Bennett1990-bc,Baym2016-kh}.  Yet the polygenic nature
of fitness makes detecting the impact selection on genome-wide variation over
short timescales in sexual populations remarkably difficult. This is because
the effect of selection on a polygenic trait (such as fitness) is distributed
across loci in proportion to their effect sizes. This can lead to subtle allele
frequency shifts on standing variation that are difficult to distinguish from
background levels of genetic drift and sampling variance.  However,
increasingly genomic experimental evolution studies with multiple timepoints,
and in some cases multiple replicate populations, are being used to detect
selected loci \parencite{Turner2011-sx,Turner2012-bm} and differentiate modes
of selection \parencite{Burke2010-tz,Barghi2019-qy,Therkildsen2019-zy}.  In
addition these temporal-genomic studies have begun in wild populations, some
with the goal of finding variants that exhibit frequency changes consistent
with fluctuating selection \parencite{Bergland2014-ij,Machado2018-cs}. In a
previous paper, we proposed that one useful signal for understanding the
genome-wide impact of polygenic linked selection detectable from temporal
genomic studies is the temporal autocovariance in allele frequency changes
\parencite{Buffalo2019-io}.  These covariances are directly estimable from
temporal genomic data and are created when the loci that underly heritable
fitness variation perturb the frequencies of linked neutral alleles; in
contrast, when genetic drift acts alone in a closed population, these
covariances are zero in expectation. Mathematically, temporal covariances are
useful because it is natural to decompose the total variance in allele
frequency change across a set of time intervals into the variances and
covariances in allele frequency change among time intervals. Furthermore,
biologically, these covariances reflect the extent to which allele frequency
changes in one generation predict changes in another due to a shared fitness
background and shared selection pressures.

Here, we provide the first empirical analyses to quantify the impact of linked
selection acting over short timescales (tens of generations) across two evolve
and re-sequence studies \parencite{Barghi2019-qy,Kelly2019-dc}, and one
artificial selection experiment \parencite{Castro2019-uk}. We repeatedly find a
signal of temporal covariance, consistent with linked selection acting to
significantly perturb genome-wide allele frequency changes across the genome in
a manner that other approaches would not be able differentiate from genetic
drift. We estimate the lower bound on the proportion of total variation in
allele frequency change caused by linked selection, and the correlation between
allele frequency changes between replicate populations caused by response to
convergent selection pressures. Overall, we demonstrate that linked selection
has a powerful role in shaping genome-wide allele frequency changes over very
short timescales.

>>>>

A long-standing problem in evolutionary genetics is quantifying the roles of
genetic drift and selection in shaping genome-wide allele frequency changes.
Selection can both directly and indirectly affect allele frequencies, with the
indirect effect coming from the action of selection on correlated loci
elsewhere in genome e.g. linked selection (\cite{Maynard_Smith1974-lc},
\cite{Charlesworth1993-gb,Nordborg1996-nq}; see \cite{Barton2000-zg} for a
review). Previous work on this question has mostly focused on teasing apart
the impacts of drift and selection on genome-wide diversity using population
samples from a single contemporary timepoint, often by modeling the correlation
between regional recombination rate, gene density, and diversity created in the
presence of linked selection \parencite{SellaReview2009,CutterPayseur}. This  %\parencite{Begun1992-ey,Elyashiv2016-vt}
approach has shown linked selection has a major role in shaping patterns of
genome-wide diversity across the genomes of \gc{a range of sexual species}
\parencite{Begun2007-bg,Beissinger2016-cm,Sattath2011-dr,Williamson2014-oy,Andersen2012-bj,Cutter2010-gi,Elyashiv2016-v},
and has allowed us to quantify the relative influence of positive selection
(hitchhiking) and negative selection (background selection;
\cite{Nordborg2005-dc,McVicker2009-ax,Hernandez2011-gs,Elyashiv2016-vt}.
However, we lack an understanding of how genome-wide linked selection acts over
time.

There are numerous examples of rapid phenotypic adaptation
\parencite{Grant2011-wk,Grant2006-hj,Reznick1997-mh,Franks2007-dr} and rapid
\gc{, selection-driven} genomic evolution in asexual populations
\parencite{Good2017-om,Bennett1990-bc,Baym2016-kh}.  Yet the polygenic nature
of fitness makes detecting the impact selection on genome-wide variation over
short timescales in sexual populations remarkably difficult. This is because
the effect of selection on a polygenic trait (such as fitness) is distributed
across loci in proportion to their effect sizes. This can lead to subtle allele
frequency shifts on standing variation that are difficult to distinguish from
background levels of genetic drift and sampling variance.  However,
increasingly genomic experimental evolution studies with multiple timepoints,
and in some cases multiple replicate populations, are being used to detect
\gc{large effect} selected loci \parencite{Turner2011-sx,Turner2012-bm} and differentiate modes
of selection \parencite{Burke2010-tz,Barghi2019-qy,Therkildsen2019-zy}.  In
addition these temporal-genomic studies have begun in wild populations, some
with the goal of finding variants that exhibit frequency changes consistent
with fluctuating selection \parencite{Bergland2014-ij,Machado2018-cs}. In a
previous paper, we proposed that one useful signal for understanding the
genome-wide impact of polygenic linked selection detectable from temporal
genomic studies is the temporal autocovariance in allele frequency changes
\parencite{Buffalo2019-io}.  These covariances are directly estimable from
temporal genomic data and are created when the loci that underly heritable
fitness variation perturb the frequencies of linked neutral alleles; in
contrast, when genetic drift acts alone in a closed population, these
covariances are zero in expectation. Mathematically, temporal covariances are
useful because it is natural to decompose the total variance in allele
frequency change across a set of time intervals into the variances and
covariances in allele frequency change among time intervals. Furthermore,
biologically, these covariances reflect the extent to which allele frequency
changes in one generation predict changes in another due to a shared \gc{genetic} %fitness
background and \gc{correlated} selection pressures. %shared

Here, we provide the first empirical analyses to quantify the impact of linked
selection acting over short timescales (tens of generations) across two evolve
and re-sequence studies \parencite{Barghi2019-qy,Kelly2019-dc}, and \gc{an}
artificial selection experiment \parencite{Castro2019-uk}. We repeatedly find a
signal of temporal covariance, consistent with linked selection acting to
significantly perturb genome-wide allele frequency changes across the genome in
a manner that other approaches would not be able differentiate from genetic
drift. We estimate the lower bound on the proportion of total variation in
allele frequency change caused by \gc{selection}, and the correlation
between  %linked
allele frequency changes between replicate populations caused by response to
convergent selection pressures. Overall, we demonstrate that linked selection
has a powerful role in shaping genome-wide allele frequency changes over very
short timescales.
